---
title: "Handwash"
author: "olayinka ogunneye"
date: "2025-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**
In the 1840s, **Dr. Ignaz Semmelweis** discovered that handwashing significantly reduced childbed fever deaths. This analysis replicates his findings using data from **Vienna General Hospital**.

### **Load Required Libraries**

```{r}
library(tidyverse)
```

### **Load and Inspect the Data**

- We will load two datasets:

- yearly_deaths_by_clinic.csv: Yearly deaths and births in two hospital clinics.

- monthly_deaths.csv: Monthly records of deaths and births.

```{r pressure, echo=FALSE}
yearly <- read_csv("C:/Users/olayi/Desktop/Data/R/R Projects/Handwashing/datasets/yearly_deaths_by_clinic.csv")

print(yearly)

monthly <- read_csv("C:/Users/olayi/Desktop/Data/R/R Projects/Handwashing/datasets/monthly_deaths.csv")
print(monthly)

```

### **Calculate Proportion of Deaths**

- We add a new column proportion_deaths to both datasets.
```{r}
yearly <- yearly %>% mutate(proportion_deaths = deaths / births)

monthly <- monthly %>% mutate(proportion_deaths = deaths / births)

```

### **Yearly Death Trends by Clinic**

A line plot showing the proportion of deaths for each clinic.
```{r}
ggplot(yearly, aes(x = year, y = proportion_deaths, color = clinic)) +
  geom_line() +
  labs(title = "Yearly Proportion of Deaths by Clinic", 
       x = "Year", 
       y = "Proportion of Deaths")

```

### **Monthly Death Trends**

A line plot showing the proportion of deaths over time.
```{r}
ggplot(monthly, aes(x = date, y = proportion_deaths)) +
  geom_line() +
  labs(title = "Monthly Proportion of Deaths", 
       x = "Year", 
       y = "Proportion of Deaths")

```

### **Effect of Handwashing**

Dr. Semmelweis introduced mandatory handwashing on June 1st, 1847. We will add a column to mark 
whether handwashing was in effect.

```{r}
handwashing_start = as.Date('1847-06-01')
monthly <- monthly %>% mutate(handwashing_started = date >= handwashing_start)

```

Now, we visualize the effect of handwashing.

```{r}
ggplot(monthly, aes(x = date, y = proportion_deaths, color = handwashing_started)) +
  geom_line() +
  labs(title = "Effect of Handwashing on Monthly Deaths", 
       x = "Year", 
       y = "Proportion of Deaths")

```

### **Impact of Handwashing on Death Rates**

We compute the average proportion of deaths before and after handwashing.

```{r}
monthly_summary <- monthly %>% 
  group_by(handwashing_started) %>%
  summarize(mean_proportion_deaths = mean(proportion_deaths, na.rm = TRUE))

monthly_summary

```

### **Conclusion**

The results confirm Dr. Semmelweis' findings: handwashing drastically reduced the death rate. 
This was a groundbreaking discovery for modern medicine!